Review: Redefining Domestication: Adaptive Introgression during Crop Expansion

### 2017-06-09 paper.tex
edits to maize and rice

The relationship between maize (\emph{Zea mays} ssp. \emph{mays}) and the teosinte \emph{Zea mays} ssp. \emph{mexicana} (hereafter referred to as \emph{mexicana}) offers a prime case study of adaptive crop/wild introgression.
Maize was domesticated from (\emph{Zea mays} ssp. \emph{parviglumis}) approximately 9,000 years ago in the Balsas River Valley in Mexico \cite{matsuoka2002single}.
From this domestication center, maize spread into the highlands of the Mexican central plateau, where it came into sympatry with wild \emph{mexicana}.
Introgression between \emph{mexicana} and Mexican highland maize has been reported based on evidence from both morphological data \cite {wilkes1977, lauter2004, doebley1984} and molecular analyses \cite{matsuoka2002, vanHeerwaarden2011, doebley1987, warburton2011, fukunaga2005}.
However, \citep{hufford2013} was the first to reveal evidence that introgression from \emph{mexicana} into Mexican highland maize is adaptive.
%The following is not fully clear.  Cut or revise.
The authors identified nine genomic regions which showed evidence of introgression from \emph{mexicana} to maize in both the HAMPMIX and the linkage model of STRUCTURE analyses with over seven sympatric population pairs among the nine pairs sampled.
Among the nine regions, three span the centromeres of chromosomes 5, 6, and 10, and one is located in the inversion polymorphism on chromosome 4, suggesting a significant role of genome structures restricting recombination in adaptive introgression.
By further characterizing the nine introgression regions, it is found that most regions contain long tracts of zero diversity, enriched with QTL linked with anthocyanin content and leaf macrohairs \cite{lauter2004} and over-represented with the SNPs demonstrating high association with temperature seasonality.
It was also revealed that in the growth chamber experiment, maize populations with introgression from \emph{mexicana} on chromosomes 4 (associated with QTLs controlling pigment density and macrohairs) and 9 (overlapped with QTLs for macrohairs) exhibited more macrohairs and greater pigmentation under the highland environmental settings than the populations with absence of introgression from \emph{mexicana}.
In summary, the introgressed alleles/haplotypes from \emph{mexicana} to maize conferred adaptation to highland habitats when maize migrated from Mexican lowlands.

Even though introgression from \emph{mexicana} to Mexican highland maize has been highly supported, it is yet unknown whether and to what extent such introgression could be found in other highland maize populations which are allopatric to \emph{mexicana}. 
A recent study \cite{Takuno2015} found little empirical or theoretical support for parallel highland adaptation in the Mexican and South American highland regions, which was explained partly by the difference in potential for adaptive introgression from wild relatives.
Adaptive SNPs in Mexican highland population were more likely located in the introgressed regions than those in South American highland populations.
Furthermore, the adaptive SNPs in the Mexican highland population were more likely also showing signatures of local adaptation in \emph{mexicana} and \emph {pulviglumis} populations than those South American highland population.
For these reasons, adaptive introgression from wild relatives may play a significant role in patterning the genetic differences of maize highland populations.

Several questions regarding \emph{mexicana} introgression yet remain.
Has the introgression from \emph{mexicana} spread to multiple highland populations? 
How do highland-adaptation traits differ among populations with and without introgression from \emph{mexicana}?
A recent study \cite{Wang2015manuscript} first utilized the ABBA-BABA statistics to evaluate the existence or absence of introgression from \emph{mexicana} to multiple highland populations, and then $\hat{f_{d}}$ statistic proposed by Martin et al. (2015) \citep{martin2015} were calculated to locate the introgressed loci. 
Figure \ref{fig:fd} shows the loess regression of $\hat{f_{d}}$ in $10kb$ non\-overlapping windows across chromosomes 4 and 5 in multiple comparisons. 
On chromosome 4 (Figure \ref{fig:fd}), Mexican highland (MexHigh) and Guatamalan highland (GuaHigh) exhibited strong evidence of introgression from \emph{mexicana}, and the peaks of distribution corresponded to the region identified in Hufford et al. (2013) \citep{hufford2013}.
The signal of introgression is absent for the other three populations.
On chromosome 5, the signals of introgression (the peak region of the distribution) are present in MexHigh, GuaHigh and Southwestern US Highland (SW\_US), but not the other two.
More details on the other chromosomes can be found in \cite{Wang2015manuscript}.
Overall, both analyses revealed that gene flow from \emph{mexicana} exists, not only to MexHigh maize, but also to GuaHigh maize, as well as to some individuals in SW\_US, but not to the Andes and South American lowland (SA\_Low) populations, suggesting different extent contributions \emph{mexicana} made to multiple maize highland landrace populations. 
The Andean maize, the population totally isolated from the occurrence of any teosinte species, underwent the severest historical bottleneck, as a population in the front wave of the serial founder effects. 
%The stronger genetic drift introduced by the smaller historical population size increased homozygosity but reduced the heterozygosity of the population, and it also extended the length of homozygosity.
The high frequency of deleterious alleles caused by stronger genetic drift in the Andes population, together with the reduced efficiency of selection against deleterious sites, contributes to the observed higher mutation load in the well-isolated maize population.
Although it is clear that the absence of introgression from wild relatives provided fewer genetic resources for highland adaptation (making the highland adaptation in the Andes unique), it is yet unknown whether or not being out of reach of wild relatives is a reason for reduced fitness in the Andes population.
Furthermore, the question of whether convergent evolution occurs between populations with and without introgression from wild relatives is a key topic for future studies in maize.

@@

Introgression appears to have played an important role in the domestication history of rice.
The center of Asian rice (\emph{Oryza sativa}) domestication is not known with complete confidence, but genetic and archaeobotanical evidence points towards both the Yangzee Basin in China and the Ganges plains in India for \emph{O. sativa}, 8,200-13,500 years ago, from wild \emph{O. rufipogon}  \cite{oka2012origin, fuller2010consilience, ricepedia, molina2011molecular}\ or "the Asian form of \emph{O. perennis} complex" \cite{oka2012origin}.
Asian rices readily hybridize both with other domesticated subspecies and with wild relatives (of which there are about twenty \cite{ricepedia}).
The high genetic diversity within \emph{O. sativa} is likely due to introgression with wild relatives both during domestication and upon the dispersal of \emph{O. sativa} into new environments and sympatry with new relatives \cite{second1982origin}.

In an attempt to categorize the varieties of Asian rice, \cite{khush2003classifying} performed isozyme polymorphism analysis (11 enzymes, 20 loci, 79 alleles) on 25,519 rice varieties.
The vast majority of the rice varieties fell into the group indica (73.45\%) or japonica (23.02\%).
Japonica rice cultivars were likely domesticated first from wild rice populations in southern China, and indica cultivars later developed through hybridization of ancient japonica with new wild rice populations in south and southeast Asia \cite{Huang2012, londo2006phylogeography}.
However, about 1.26\% belonged to four other groups (and 2.27\% were indeterminate).
Some of the varieties from these other four groups most likely arose from hybrids of cultivated rice and \emph{O. rufipogon}.
Introgression of alleles between cultivated rices, as well as between cultivated and wild rices, is common \cite{oka2012origin, second1982origin, zhao2010genomic}\, and natural introgression from wild to domesticate is suspected (\cite{zhao2010genomic} calls for research into this possibility).
These rices often have blight resistance gene, likley imparted from the wild relative parent.
Several resistance genes (grassy stunt virus, bacterial blight, brown planthopper, blast) are known to have been introgressed from wild relatives into \emph{O. sativa} by researchers \cite{brar1997alien, khush1974inheritance}.
Some aus and rayada varieties retain characteristics of the putative wild parent.
Aus exhibits a sprawling growth pattern and easy-threshing grain.
Rayada is adapted to areas with longer periods of flooding, with heightened seed dormancy and photoperiod sensitivity (permitting harvest during times without standing water).
Aswina variety, adapted to growth in deepwater conditions with a short period of flooding, seems to be the result of hybridization between \emph{O. sativa} and deepwater-adapted \emph{O. rufipogon} populations.

In addition to investigative experiments, gene flow from wild relatives has been utilized to produce agronomic rice varieties.
Yatsen No. 1, for example, showed resistance to pests and diseases and adapted well to environmental conditions \cite{ting1933wild}.
Several lines were derived from Yatsen No. 1, and went on to be utilized extensively in parts of China.

@@

Domesticated and wild barleys belong to the same species, \emph{Hordeum vulgare}, and are capable of producing viable offpsring via hybridizaion \cite{von1995ecographical}.
Barley (\emph{Hordeum vulgare} subsp. \emph{vulgare}) is believed to have been domesticated at least twice, once from wild subsp. \emph{spontaneum} in the Fertile Crescent and once from subsp. \emph{spontaneum} var. \emph{agriocrithon} in Tibet roughly 10,000 years ago \cite{takahashi1955origin, badr2000origin, oka2012origin, azhaguvel2007phylogenetic, haberer2015barley}\.
However, many of the details of barley domestication are still disputed.

>>>>

The relationship between maize (\emph{Zea mays} ssp. \emph{mays}) and the teosinte \emph{Zea mays} ssp. \emph{mexicana} (hereafter referred to as \emph{mexicana}) offers a prime case study of adaptive wild-to-crop introgression.
Maize was domesticated from (\emph{Zea mays} ssp. \emph{parviglumis}) approximately 9,000 years ago in the lowlands of the Balsas River Valley in Mexico \cite{matsuoka2002single}.
From this domestication center, maize spread into the highlands of the Mexican Central Plateau, where it came into sympatry with wild \emph{mexicana}.
Introgression from \emph{mexicana} to maize in the highlands of Mexico has been reported based on evidence from both morphological data \cite {wilkes1977, lauter2004, doebley1984} and molecular analyses \cite{matsuoka2002, vanHeerwaarden2011, doebley1987, warburton2011, fukunaga2005}.
However, \citep{hufford2013} first localized \emph{mexicana} introgression to chromosomal regions and provided evidence that it was likely adaptive.
%The following is not fully clear.  Cut or revise.
The authors identified nine genomic regions in several maize populations which showed evidence of\emph{mexicana} introgression based on local ancestry deconvolution using HAPMIX and results from the linkage model of STRUCTURE.
These introgressed segments showed low diversity and overlapped QTL that had previously been found to control anthocyanin content and leaf macrohairs \cite{lauter2004}, traits known to be important in adaptation to high elevation.
In a growth chamber experiment, the authors demonstrated that maize populations with \emph{mexicana} introgression showed greater plant height (a proxy for fitness) under highland environmental settings than populations that lacked introgression.
Height differences were not detected under lowland conditions.
%Among the nine regions, three span the centromeres of chromosomes 5, 6, and 10, and one is located in the inversion polymorphism on chromosome 4, suggesting a significant role of genome structures restricting recombination in adaptive introgression.
%In summary, the introgressed alleles/haplotypes from \emph{mexicana} to maize conferred adaptation to highland habitats when maize migrated from Mexican lowlands.

%Even though introgression from \emph{mexicana} to Mexican highland maize has been highly supported, it is yet unknown whether and to what extent such introgression could be found in other highland maize populations which are allopatric to \emph{mexicana}. 
%A recent study \cite{Takuno2015} found little empirical or theoretical support for parallel highland adaptation in the Mexican and South American highland regions, which was explained partly by the difference in potential for adaptive introgression from wild relatives.
%Adaptive SNPs in Mexican highland population were more likely located in the introgressed regions than those in South American highland populations.
%Furthermore, the adaptive SNPs in the Mexican highland population were more likely also showing signatures of local adaptation in \emph{mexicana} and \emph {pulviglumis} populations than those South American highland population.
%For these reasons, adaptive introgression from wild relatives may play a significant role in patterning the genetic differences of maize highland populations.

Populations of \emph{mexicana} cannot be found outside of Mexico, yet maize has colonized and adapted to high elevation in a number of additional regions.
%Several questions regarding \emph{mexicana} introgression yet remain.
%Has the introgression from \emph{mexicana} spread to multiple highland populations? 
%How do highland-adaptation traits differ among populations with and without introgression from \emph{mexicana}?
A recent study \cite{Wang2015manuscript} employed the ABBA-BABA and $\hat{f_{d}}$ statistics to evaluate whether maize with \emph{mexicana} introgression was transferred to other highland regions or whether highland adaptation was obtained \emph{de novo} outside of Mexico (Figure \ref{fig:fd}).
%On chromosome 4 (Figure \ref{fig:fd}), Mexican highland (MexHigh) and Guatamalan highland (GuaHigh) exhibited strong evidence of introgression from \emph{mexicana}, and the peaks of distribution corresponded to the region identified in Hufford et al. (2013) \citep{hufford2013}.
%The signal of introgression is absent for the other three populations.
%On chromosome 5, the signals of introgression (the peak region of the distribution) are present in MexHigh, GuaHigh and Southwestern US Highland (SW\_US), but not the other two.
%More details on the other chromosomes can be found in \cite{Wang2015manuscript}.
Overall, analyses revealed that maize landraces with \emph{mexicana} introgression were transferred to nearby high elevation regions in Guatemala and the southwestern United States, but that more distant high elevation regions (\emph{e.g.,} the Andes) showed no \emph{mexicana} ancestry. \mbh{on the fence about including deleterious allele results from introgressed regions}
%The Andean maize, the population totally isolated from the occurrence of any teosinte species, underwent the severest historical bottleneck, as a population in the front wave of the serial founder effects. 
%The stronger genetic drift introduced by the smaller historical population size increased homozygosity but reduced the heterozygosity of the population, and it also extended the length of homozygosity.
%The high frequency of deleterious alleles caused by stronger genetic drift in the Andes population, together with the reduced efficiency of selection against deleterious sites, contributes to the observed higher mutation load in the well-isolated maize population.
%Although it is clear that the absence of introgression from wild relatives provided fewer genetic resources for highland adaptation (making the highland adaptation in the Andes unique), it is yet unknown whether or not being out of reach of wild relatives is a reason for reduced fitness in the Andes population.
%Furthermore, the question of whether convergent evolution occurs between populations with and without introgression from wild relatives is a key topic for future studies in maize.

@@

Introgression appears to have played an important role in the domestication history of rice.
The center of Asian rice (\emph{Oryza sativa}) domestication is not known with complete confidence, but genetic and archaeobotanical evidence point toward both the Yangzee Basin in China and the Ganges plains in India. \mbh{there is a serious debate over whether there were one or two domestication centers of rice...I'd be sure to read up on the most current thoughts about this so that we don't upset potential reviewers}
Domestication occurred 8,200-13,500 BP from the wild species \emph{O. rufipogon} \cite{oka2012origin, fuller2010consilience, ricepedia, molina2011molecular}. \mbh{here limit to one or two citations}
Asian rice readily hybridizes with other domesticated subspecies and with wild relatives (of which there are about twenty \cite{ricepedia}). \mbh{probably best to find a primary literature citation for this (journal article rather than ricepedia).  What other domesticated subspecies are referred to here?  Also, when referring to Asian rice, is this both japonica and indica?  Finally, what are the twenty relatives?  My understanding is gene flow has mostly been with rufipogon and nivara}
The high genetic diversity within \emph{O. sativa} is likely due to introgression from wild relatives both within the domestication center(s) and in new environments where rice has dispersed following domestication \cite{second1982origin}.

Asian rice can be broadly clustered into two groups, japonica and indica \cite{khush2003classifying}.
%The vast majority of the rice varieties fell into the group indica (73.45\%) or japonica (23.02\%).
Japonica rice cultivars were likely domesticated first from wild rice populations in southern China, and indica cultivars later developed through hybridization of ancient japonica with new wild rice populations in south and southeast Asia \cite{Huang2012, londo2006phylogeography}. \mbh{Is this thought to be due to adaptation?  Did the wild rice populations confer local adaptation? If so, I'd state this and explain what type of adaptation this might have been if it's known}
%However, about 1.26\% belonged to four other groups (and 2.27\% were indeterminate).
%Some of the varieties from these other four groups most likely arose from hybrids of cultivated rice and \emph{O. rufipogon}.
Introgression of alleles between cultivated rice populations, as well as between cultivated and wild rice, is common \cite{oka2012origin, second1982origin, zhao2010genomic}\, and natural introgression from wild to domesticated rice is suspected (\cite{zhao2010genomic} calls for research into this possibility). \mbh{strange sentence here...says introgression between cultivated and wild rice is common but then says introgression from wild to cultivated is just suspected...need to clarify}
%These rices often have blight resistance gene, likely imparted from the wild relative parent.
Several resistance genes (grassy stunt virus, bacterial blight, brown planthopper, blast) are known to have been introgressed from wild relatives into \emph{O. sativa} by researchers \cite{brar1997alien, khush1974inheritance}. \mbh{don't think this is what we're after in these case studies.  We want to narrow our focus to natural introgression from wild to crop species during the expansion of the crop. Wild introgression during breeding is interesting, but not the focus of the paper}.
Some aus and rayada varieties retain characteristics of the putative wild parent. \mbh{more background on aus, rayada and aswina...what is there origin? how doe they relate to japonica and indica?}
Aus exhibits a sprawling growth pattern and easy-threshing grain.
Rayada is adapted to areas with longer periods of flooding, with heightened seed dormancy and photoperiod sensitivity (permitting harvest during times without standing water).
Aswina variety, adapted to growth in deepwater conditions with a short period of flooding, seems to be the result of hybridization between \emph{O. sativa} and deepwater-adapted \emph{O. rufipogon} populations. \mbh{need citations regarding these varieties}

In addition to investigative experiments, gene flow from wild relatives has been utilized to produce agronomic rice varieties.
Yatsen No. 1, for example, showed resistance to pests and diseases and adapted well to environmental conditions \cite{ting1933wild}.
Several lines were derived from Yatsen No. 1, and went on to be utilized extensively in parts of China. \mbh{again, don't think this is what we're after with these case studies}

@@

Domesticated and wild barley belong to the same species, \emph{Hordeum vulgare}, and are capable of producing viable offpspring via hybridization \cite{von1995ecographical}.
Barley (\emph{Hordeum vulgare} subsp. \emph{vulgare}) is believed to have been domesticated at least twice roughly 10,000 BP, once from wild subsp. \emph{spontaneum} in the Fertile Crescent and once from subsp. \emph{spontaneum} var. \emph{agriocrithon} in Tibet \cite{takahashi1955origin, badr2000origin, oka2012origin, azhaguvel2007phylogenetic, haberer2015barley}.
However, details of barley domestication are still disputed.
